### 2019-10-14 main_text/main_text.tex
Conclusions edits and more

\todo{MJM: Is this para \& next excessively detailed for main text?
maybe replace w/ 1-2 sentences about low affinity Mich-Ment transporters
that supplement XapB, \& shift all this to supplement?
If we want to emphasize the general class of model rather than the particular
biology details of Xap}
Transport of xanthosine across the cell membrane happens in two steps which
we will discuss now. All statements in this paragraph are based on data
from~\cite{Norholm2001}. In a first transport step, xanthosine passes the
outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
be a rate-limiting step. Then, XapB actively transports it across the inner
membrane, powered by the proton motive force. There are two other active
nucleoside transporters that seem to be transporting xanthosine with a very
low affinity: NupC and NupG. It was found that \textDelta\emph{nupC}
\textDelta\emph{nupG} strains cannot grow on xanthosine. Hence, these seem
to be necessary to ``start'' the system by importing small amounts of
xanthosine that then activate \emph{xapB}, and there appears to be no other
significant way in which xanthosine can enter the cell.

@@

The results agree very well with the deterministic fixed points and
experiments: the mean numbers of mRNA, protein, and xanthosine in the
stochastic results are as predicted from the phase portraits. It does,
however, become clear that the phase portraits do not tell whether the cells
will actually populate both the high and the low expression state: as can be
seen in Fig~\subref{fig8:stochC}{(A)}, the cells did not switch to the high
expression state for the waiting time in our simulations, even though one
would naively expect this from the phase portrait.

@@

Note that while the deterministic analysis assumes the variables to be
continuous, the simulations work with discrete numbers of mRNA, protein, and
xanthosine. This per se is no problem, because the deterministic analysis
describes the mean values and the simulation fluctuates around this mean.
If, however, the mRNA number at the third, high fixed point is too low, the
system will not be able to resolve the two points anymore. This becomes a
problem when the distance between the first and the third fixed point
becomes as low as the stochastic fluctuations in the system, which is around
3 mRNA molecules. In Fig~\ref{fig8:stochC}, the mRNA number at the low fixed
point was fluctuating between 0 and 2 with a mean significantly smaller than
1, and at the high fixed point it was varying between 2 and 6.

@@

Comparing this to the experimentally expected timescales is difficult,
because the switching time strongly depends on the extracellular xanthosine
concentration. Experiments were always stopped after a few hours, and in
this time, the cell population might not reach its steady-state expression
distribution. Hence, the distribution could be bimodal when the experiment
is stopped but become unimodal after further waiting. That way,
extracellular xanthosine concentrations that are too high for deterministic
bistability could lead to experimental bimodality if the experiment is
stopped too early. In this case, the observed switching time would be
shorter, which makes the comparison to our simulations even harder. Thus, we
cannot say if it is problematic that the $10^5 \unit{s}$ is larger than what
was found in the experiment.

@@

\paragraph*{Hysteresis.} 
In the previous two paragraphs, the simulations were started at initial
intracellular concentrations of 0 to investigate what happens if xanthosine
is suddenly added to the cell's environment. We can now ask the opposite
question: what happens when xanthosine is removed from the extracellular
environment? To answer this, the simulation was started with initially fully
induced cells, i.e. at the mRNA, protein and intracellular xanthosine counts
of the high fixed point in the corresponding phase portrait.

The resulting distributions can be found in~\nameref{S1_Text}. They show
clear hysteresis effects: at concentrations where all cells remained in the
low expression state when being started there before, all cells now remain
at high expression. Only when the second and the third fixed point are very
close does switching to the uninduced state happen. Interestingly, this
behavior is symmetric to the ``switching on'' in the previous paragraphs,
where the lower two fixed points needed to be about as close as the upper
two need to be now for ``switching off''. The extracellular xanthosine
concentration (for the other parameters being as in
Table~\ref{table1:nondim}) needed to be as low as $\n{[c]_a} = 10$ to
observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression
state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
10^{-5} \unit{M}$). \todo{Is this understandable without seeing the data? Do
you have any suggestions to make it clearer?}

We have learned that cells only change their metabolism to xanthosine if
enough of the latter is around, but after they have switched, this metabolic
state is stable even if the xanthosine concentration decreases to a certain
extent. This explains what has been observed by Novick and
Weiner~\cite{Novick1957} for the \emph{lac} operon: when induced cells were
transferred into lower concentrations of lactose, they remained induced,
even though uninduced cells could not become induced at these
concentrations.

\section*{Conclusion}
\paragraph*{A genetic switch for metabolic adaptation with remarkable features.}
In this paper, we propose a model for biological pathway systems containing
a membrane transporter whose gene expression is, directly or indirectly,
activated by its substrate. We have shown that such a system can be bistable
and thus work as a genetic switch which reacts to the extracellular
concentration of the relevant metabolite. This switch has very useful
biological features: first, coupling of the transporter with, for example, a
degradation enzyme creates a genetic switch that enables short-term
adaptation of the cell's metabolism to its environment. Second, the switch
is stabilized by hysteresis effects when the extracellular xanthosine
concentration decreases, which explains previous experimental findings.

@@

\paragraph*{Possibilities for further investigation.}
Phase diagrams, showing for which parameters the system is bistable and for
which there is only the lower or the upper stable fixed point, could be
calculated from arguments made in~\cite{Cherry2000}. However, the
simulations have shown that deterministic bistability does not mean that
bimodality occurs, which is why we have refrained from showing such
diagrams. Furthermore, the timescales in the problem could be investigated
more thoroughly, for example the dependence of the switching time on
$\n{[c]_a}$. Having said that, the simulation for such an analysis should
account for the burstiness of transcription and translation as well as for
cell divisions, which is anything but straightforward. Lastly, some rigorous
work could be done to identify the most and least relevant parameters of the
system to create a simpler, minimal model that captures the most relevant
features.

\paragraph*{General lessons learned for the modeling of other systems.}
We point out again that assuming the mRNA concentration to be at
steady-state can have a significant impact on the apparent dynamics of a
system, as we have seen in the 3-dimensional phase portraits. Moreover, we
remind the reader that deterministic phase portraits do not show which
states are actually populated, and thus hide essential effects like
hysteresis. Lastly, we have exemplified issues with the common picture of
measuring cooperativity in terms of the Hill coefficient: the Hill
coefficient simply denotes the number of binding sites and does not reflect
how cooperative these binding sites are.

\paragraph*{A successful model.}
All model parameters could be estimated fairly well despite the lack of deep
experimental knowledge about the model system. In the results, every
variable and parameter has its expected order of magnitude, which suggests
that the model captures all relevant features and is able to describe the
dynamics of the \emph{xap} system. With the framework given in this text, it
should be straightforward to model other promoters, regulatory pathways or
enzymes and thereby adapt the model to other genes and metabolites. Examples
include \emph{lac}, \emph{ara}, and \emph{xyl}, but we suspect that many if
not most metabolic processes involve the adaptation mechanism that we have
investigated here, and that a lot can be understood about them through our
model. This apparent success demonstrates once more that even for broadly
unknown systems, rigorous physical modeling can potentially offer an
efficient way to gain a very thorough understanding of the behavior of the
system.

\todo{Is the conclusion too long?}

@@

\section*{Acknowledgments}
We thank Griffin Chure, Charlotte Strandkvist, and Kyle Naughton for
providing their data on the \emph{xap} system and pointing out the
bistability. We also thank Griffin Chure for answering any biological
questions that arose, Nathan Belliveau for bringing similar systems to our
attention, and Jane Kondev and Jin Wang for interesting discussions. Plots
were done in \emph{Mathematica} using the packages MaTeX and
CurvesGraphics6. K.S.L. acknowledges financial support from the
Werner~Siemens-Foundation through the Swiss Study Foundation.
\todo[inline]{I don't really know what/who is usually listed here... Did I list too many people (I don't know at what ``extent'' of help people are usually named here) or did I forget anyone? Should I list everyone who worked on the xap system last summer and provided the data even though they are already cited, but only as Griffin et al.? Is it enough to list packages like this or should I try to find some website to cite? Should I list Python \& StochPy again it's already mentioned in the text? SHould I thank someone for the providing the elements for the figures and whom? Is there funding that should be listed?}

@@

\todo[inline]{Should I also list the references that are cited in the appendix?}

\todo[inline]{The capitalization style is different for the titles of the references. Should I list them as they are on the corresponding paper or should I do it uniformly? Does it matter at all? Otherwise I leave it as it is.}

>>>>

\todo[inline]{MJM: Are the following 2 paragraphs excessively detailed for main text?
maybe replace w/ 1-2 sentences about low affinity Mich-Ment transporters
that supplement XapB, \& shift all this to supplement?
If we want to emphasize the general class of model rather than the particular
biology details of Xap}
Transport of xanthosine across the cell membrane happens in two steps which
we will discuss now. All statements in this paragraph are based on data
from~\cite{Norholm2001}. In a first transport step, xanthosine passes the
outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
be a rate-limiting step. Then, XapB actively transports it across the inner
membrane, powered by the proton motive force. There are two other active
nucleoside transporters that seem to be transporting xanthosine with a very
low affinity: NupC and NupG. It was found that \textDelta\emph{nupC}
\textDelta\emph{nupG} strains cannot grow on xanthosine. Hence, these seem
to be necessary to ``start'' the system by importing small amounts of
xanthosine that then activate \emph{xapB}, and there appears to be no other
significant way in which xanthosine can enter the cell.

@@

The results agree very well with the deterministic fixed points and
experiments: the mean numbers of mRNA, protein, and xanthosine in the
stochastic results are as predicted from the phase portraits. It does,
however, become clear that the phase portraits do not tell whether the cells
will actually populate both the high and the low expression state: as can be
seen in Fig~\subref{fig8:stochC}{(A)}, the cells never switched to the high
expression state during the run time of our simulations.
\todo{MJM: depends on an effective barrier height, hard to intuit from phase port.
Bistability rather than bimodality means it may very well practically never switch.}

@@

Note that while the deterministic analysis assumes the variables to be
continuous, the simulations work with discrete numbers of mRNA, protein, and
xanthosine. This per se is no problem, because the deterministic analysis
describes the mean values and the simulation fluctuates around this mean.
If, however, the mRNA number at the third (high) fixed point is too low, the
system will not be able to resolve the two points anymore. This becomes a
problem when the distance between the first and the third fixed point
becomes as low as the stochastic fluctuations in the system, which is around
3 mRNA molecules. In Fig~\ref{fig8:stochC}, the mRNA number at the low fixed
point was fluctuating between 0 and 2 with a mean significantly smaller than
1, and at the high fixed point it was varying between 2 and 6.

@@

Comparing this to the experimentally expected timescales~\cite{Chure2019} is difficult,
because the switching time strongly depends on the extracellular xanthosine
concentration. Experiments were always stopped after a few hours, and in
this time, the cell population might not reach its steady-state expression
distribution. Hence, the distribution could be bimodal when the experiment
is stopped but become unimodal after further waiting. That way,
extracellular xanthosine concentrations that are too high for deterministic
bistability could lead to experimental bimodality if the experiment is
stopped too early. In this case, the observed switching time would be
shorter, which makes the comparison to our simulations even harder. Thus, we
cannot say if it is problematic that the $10^5 \unit{s}$ is larger than what
was found in the experiment.

@@

\paragraph*{Hysteresis.} 
In Figs~\ref{fig8:stochC} and~\ref{fig:StochT}, the simulations were started at initial
intracellular concentrations of 0 to investigate what happens if xanthosine
is suddenly added to the cell's environment. We can now ask the opposite
question: what happens when xanthosine is removed from the extracellular
environment? To answer this, the simulation was started with initially fully
induced cells, i.e. at the mRNA, protein and intracellular xanthosine counts
of the high fixed point in the corresponding phase portrait.

The resulting distributions can be found in~\nameref{S1_Text}. They show
clear hysteresis effects: there exist extracellular xanthosine concentrations where
initially uninduced cells remained uninduced and initially induced cells remain induced.
Only when the second and the third fixed point are very close can initially
induced cells switch to the uninduced state. Interestingly, this
behavior is symmetric to the ``switching on'' in the previous paragraphs,
where the lower two fixed points needed to be about as close as the upper
two need to be now for ``switching off''. %The extracellular xanthosine
% concentration (for the other parameters being as in
% Table~\ref{table1:nondim}) needed to be as low as $\n{[c]_a} = 10$ to
% observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression
% state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
% 10^{-5} \unit{M}$).
\todo[inline]{Is this understandable without seeing the data? Do
you have any suggestions to make it clearer?
MJM: I think the quantitative details can go to the supplement where they
have more context, the qualitative observation of hysteresis is enough here.}

In other words, cells only change their metabolism to xanthosine if
enough of the latter is around, but after they have switched, this metabolic
state is stable even if the xanthosine concentration decreases to a certain
extent. This explains what has been observed by Novick and
Weiner~\cite{Novick1957} for the \emph{lac} operon: when induced cells were
transferred into lower concentrations of lactose, they remained induced,
even though uninduced cells could not become induced at these
concentrations.

\section*{Conclusion}
\paragraph*{A genetic switch for metabolic adaptation with remarkable features.}
In this paper, we propose a model for genetic circuits containing
a membrane transporter whose gene expression is, directly or indirectly,
activated by its substrate. We have shown that such a system can be bistable
and thus work as a genetic switch which reacts to the extracellular
concentration of the relevant metabolite. This switch has very useful
biological features. First, coupling of the transporter with, for example, an
enzyme which metabolizes the transporter substrate creates a genetic switch that enables short-term
adaptation of the cell's metabolism to its environment. Second, the switch
is stabilized by hysteresis effects when the extracellular substrate
concentration decreases, which explains previous experimental findings.

@@

\paragraph*{Possibilities for further investigation.}
Phase diagrams, showing for which parameters the system is bistable and for
which there is only the lower or the upper stable fixed point, could be
calculated from arguments made in~\cite{Cherry2000}. However, the
simulations have shown that deterministic bistability does not mean that
bimodality occurs, which is why we have refrained from showing such
diagrams. Furthermore, the timescales in the problem could be investigated
more thoroughly, for example the dependence of the switching time on
$\n{[c]_a}$, but such an analysis would probably need to
account for the burstiness of transcription and translation as well as for
cell divisions, which is anything but straightforward. Lastly, some rigorous
work could be done to identify the most and least relevant parameters of the
system to create a simpler, minimal model that captures the most relevant
features.
\todo[inline]{MJM: I think this last sentence sells us short: qualitative changes result
if we remove any present feature of the model.
That sounds to me like a distillation of the essential.}

\paragraph*{General lessons learned for the modeling of other systems.}
We point out again that assuming the mRNA concentration to be at
steady-state can have a significant impact on the apparent dynamics of a
system, as we have seen in the 3-dimensional phase portraits. Moreover, we
remind the reader that deterministic phase portraits do not show which
states are actually populated, and thus hide essential effects like
hysteresis. Lastly, we have exemplified issues with the common picture of
measuring cooperativity in terms of the Hill coefficient: the Hill
coefficient simply denotes the number of binding sites and does not reflect
how cooperative these binding sites are. 
\todo[inline]{MJM: I'm not sure we've really proven the asserted points re 3D \& Hill.
I would advocate removing this whole paragraph.}

\paragraph*{A successful model.}
All model parameters could be reasonably estimated despite the paucity of
experimental knowledge about the model system. In the results, every
variable and parameter \todo{MJM: I'm confused,
we estimated all the model parameters, so what expectations are we comparing to?}
has its expected order of magnitude, which suggests
that the model captures all relevant features and is able to describe the
dynamics of the \emph{xap} system. With the framework given in this text, it
should be straightforward to model other promoters, regulatory pathways or
enzymes and thereby adapt the model to other genes and metabolites. Examples
include \emph{lac}, \emph{ara}, and \emph{xyl}, but we suspect that many if
not most metabolic processes involve the adaptation mechanism that we have
investigated here, and that much can be understood about them through our
model. This apparent success demonstrates once more that even for broadly
unknown systems, rigorous physical modeling can potentially offer an
efficient way to gain a very thorough understanding of the behavior of the
system.

\todo{Is the conclusion too long? MJM: not necessarily}

@@

\section*{Acknowledgments}
We thank Griffin Chure, Charlotte Strandkvist, and Kyle Naughton for
providing their data on the \emph{xap} system and pointing out the
bistability. We also thank Griffin Chure for answering the many biological
questions that arose, Nathan Belliveau for bringing similar systems to our
attention, and Jane Kondev and Jin Wang for interesting discussions.
For assistance with illustrations, we also thank Nigel Orme and Griffin Chure.
\todo{MJM: this kinda reads like GC should be in the author list...}
K.S.L. acknowledges financial support from the
Werner~Siemens-Foundation through the Swiss Study Foundation.
This material is based upon work supported by the National Science Foundation
Graduate Research Fellowship under Grant No.\ DGE-114446.
\todo{MJM: need to double-check my NSF \#}
\todo[inline]{MJM: Software/code acknowledgements usually go in a separate section,
and if we want we could set up a code repo to cite along with that.}

@@

\todo[inline]{Should I also list the references that are cited in the appendix? MJM: no, I think the supplement has its own separate list.}

\todo[inline]{The capitalization style is different for the titles of the references. Should I list them as they are on the corresponding paper or should I do it uniformly? Does it matter at all? Otherwise I leave it as it is. MJM: I think the copy editors will sort it out.}
